---
# Documentation: https://sourcethemes.com/academic/docs/managing-content/

title: "Residential Zoning in Pittsburgh"
subtitle: ""
summary: ""
authors: [Conor Tompkins]
tags: [R, Pittsburgh]
categories: [R, Pittsburgh]
date: 2019-06-22
lastmod: 2020-08-28T09:36:42-04:00
featured: false
draft: false

# Featured image
# To use, add an image named `featured.jpg/png` to your page's folder.
# Focal points: Smart, Center, TopLeft, Top, TopRight, Left, Right, BottomLeft, Bottom, BottomRight.
image:
  caption: ""
  focal_point: ""
  preview_only: false

# Projects (optional).
#   Associate this post with one or more of your projects.
#   Simply enter your project's folder or file name without extension.
#   E.g. `projects = ["internal-project"]` references `content/project/deep-learning/index.md`.
#   Otherwise, set `projects = []`.
projects: []
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE)
```

The [New York Times recently published an article](https://www.nytimes.com/interactive/2019/06/18/upshot/cities-across-america-question-single-family-zoning.html) about zoning in U.S. cities, particularly single-unit detached residential housing. The article did not include Pittsburgh, so I downloaded the [zone shapefile from the WPRDC](https://data.wprdc.org/dataset/zoning1) and made my own map.

This blog quickly goes through the steps to make the map and other graphs about the data.

First, load the required libraries and set up the environment:
```{r}
library(tidyverse)
library(sf)
library(ggmap)
library(janitor)
library(hrbrthemes)

options(scipen = 999)
```

Read in the shapefile with `st_read` and inspect the data with `glimpse`:
```{r}
shapefile <- st_read("data/Zoning-shp/Zoning.shp")

glimpse(shapefile)
```

We need to munge the data to get it in shape for analysis. This makes some simple TRUE|FALSE flags for basic zone information and uses `case_when` to create `type`, which represents aggregated zone types.
```{r}
df <- shapefile %>% 
  mutate(residential = str_detect(full_zonin, "RESIDENT"),
         single_unit = str_detect(full_zonin, "SINGLE-UNIT"),
         attached = str_detect(full_zonin, "ATTACHED"),
         type = case_when(residential == TRUE & single_unit == TRUE & attached == FALSE ~ "Single-unit detached residential",
                          residential == TRUE & single_unit == FALSE | attached == TRUE ~ "Other residential",
                          full_zonin == "EDUCATIONAL/MEDICAL INSTITUTION" ~ "Educational/Medical",
                          residential == FALSE ~ "Other non-residential"),
         type = factor(type, levels = c("Single-unit detached residential", 
                                        "Other residential",
                                        "Educational/Medical",
                                        "Other non-residential")),
         alpha_flag = type == "Single-unit detached residential")
```

```{r, echo = FALSE}
df %>% 
  st_drop_geometry() %>% 
  glimpse()
```

This counts the number of rows per full zone description (`full_zonin`) and `type`:
```{r}
df_zones <- df %>% 
  count(full_zonin, type, sort = TRUE) %>% 
  st_drop_geometry()
```

```{r, echo = FALSE}
glimpse(df_zones)
```

Create a basic bar chart to show the distribution of `type`:
```{r fig.height=4, fig.width=9}
df %>% 
  st_drop_geometry() %>% 
  group_by(type, residential) %>% 
  summarize(area = sum(area)) %>% 
  ungroup() %>% 
  mutate(type = fct_reorder(type, area)) %>% 
  ggplot(aes(type, area / 1000000, fill = residential)) +
    geom_col() +
    scale_y_comma() +
    scale_fill_discrete("Is the zone residential?") +
    labs(x = "Zone type",
         y = "Land area in millions of feet") +
    coord_flip() +
    theme_ipsum()
```

Use a bar chart to show the distribution of `full_zonin`:
```{r fig.height=12, fig.width=12}
df %>% 
  st_drop_geometry() %>% 
  group_by(full_zonin, residential) %>% 
  summarize(area = sum(area)) %>% 
  ungroup() %>% 
  mutate(full_zonin = fct_reorder(full_zonin, area)) %>% 
  ggplot(aes(full_zonin, area / 1000000, fill = residential)) +
    geom_col() +
    scale_y_comma() +
    scale_fill_discrete("Is the zone residential?") +
    labs(x = "Full zone description",
         y = "Land area in millions of feet") +
    coord_flip() +
    theme_ipsum()
```

This calculates the total land area zoned for any type of residential housing:
```{r}
df %>% 
  st_drop_geometry() %>% 
  mutate(single_unit_flag = type == "Single-unit detached residential") %>% 
  filter(residential == TRUE) %>% 
  summarize(total_area = sum(area))
```

This calculates the % of residential zoning that is zoned for single-unit detached residential housing units:
```{r}
df %>% 
  st_drop_geometry() %>% 
  filter(residential == TRUE) %>% 
  mutate(single_unit_flag = (type == "Single-unit detached residential")) %>% 
  group_by(single_unit_flag) %>% 
  summarize(zone_area = sum(area)) %>% 
  mutate(pct_area = zone_area / sum(zone_area))
```

This creates a map of the zones, fills them by `type`, and overlays it on a GoogleMaps basemap. I also insert the boundaries of the City of Pittsburgh.
```{r fig.height=12, fig.width=12}
city_boundary <- st_read("data/Pittsburgh_City_Boundary-shp/City_Boundary.shp")

#register_google(key = "your key here")
pgh_map <- get_map(location = c(lat = 40.445315, lon = -79.977104), 
                   zoom = 12, maptype = "roadmap", source = "google")


ggmap(pgh_map) +
  geom_sf(data = df %>% filter(type != "Other non-residential"), aes(fill = type), inherit.aes = FALSE, size = .5, alpha = 1, color = NA) +
  geom_sf(data = city_boundary, inherit.aes = FALSE, alpha = 0, size = 2) +
  coord_sf(crs = st_crs(4326)) +
  scale_fill_manual("Zone type",
                      values = c("#ea60b9", "#4cafc5", "yellow", "light grey")) +
  labs(title = "56% of residential zoned land area is single-family detached residential",
         subtitle = "City of Pittsburgh zoning",
         caption = "@conor_tompkins, data from WPRDC") +
  theme_void()
```

I used `scale_fill_manual` to manually set the color palette to match the NYTimes article.