---
# Documentation: https://sourcethemes.com/academic/docs/managing-content/

title: "Effect of Geographic Resolution on ebirdst Abundance"
subtitle: ""
summary: ""
authors: [Conor Tompkins]
tags: [R, eBird]
categories: [R, eBird]
date: 2021-11-23T08:20:43-05:00
lastmod: 2021-11-23T08:20:43-05:00
featured: false
draft: false

# Featured image
# To use, add an image named `featured.jpg/png` to your page's folder.
# Focal points: Smart, Center, TopLeft, Top, TopRight, Left, Right, BottomLeft, Bottom, BottomRight.
image:
  caption: ""
  focal_point: ""
  preview_only: false

# Projects (optional).
#   Associate this post with one or more of your projects.
#   Simply enter your project's folder or file name without extension.
#   E.g. `projects = ["internal-project"]` references `content/project/deep-learning/index.md`.
#   Otherwise, set `projects = []`.
projects: []
---

---
title: "Effect of geographic resolution on ebirdst abundance"
author: "Conor Tompkins"
date: "11/23/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE)
```

While exploring some of the citizen science bird observation data available through [`ebirdst`](https://cornelllabofornithology.github.io/ebirdst/), I was confused by how to understand the calculation of `ebirdst`'s `abundance` metric.

The `ebirdst` documentation (`?ebirdst::load_raster`) defines `abundance` as:

>the expected relative abundance, computed as the product of the probability of occurrence and the count conditional on occurrence, of the species on an eBird Traveling Count by a skilled eBirder starting at the optimal time of day with the optimal search duration and distance that maximizes detection of that species in a region.

I had seen some weird results when trying to manually calculate `abundance` as `occurrence * count`. My initial attempt had aggregated the results by month.

The underlying problem is that `abundance` and `count` are the results of models, and are subject to model error. I also believe that the data outputted from `load_raster` lacks the necessary significant digits to accurately recreate `abundance`. Lowering the resolution or aggregating the data will exacerbate this issue.

This code loads my [convenience function](https://github.com/conorotompkins/ebird_shiny_app/blob/main/scripts/functions/pull_species_metric.R) to retrieve a metric for a species at a given geographic resolution. This gets `occurrence`, `count`, and `abundance` for the Northern Cardinal at high (3 km), medium (9 km), and low resolutions (27 km). The function also crops the underlying raster data to Allegheny County, PA.
```{r}
library(here)
library(hrbrthemes)
library(patchwork)

source("https://raw.githubusercontent.com/conorotompkins/ebird_shiny_app/main/scripts/functions/pull_species_metric.R")

theme_set(theme_ipsum())

species_table <- crossing(species = c("Northern Cardinal"),
                          metric = c("occurrence", "count", "abundance"),
                          resolution = c("hr", "mr", "lr"))

species_table
```

```{r cache=TRUE}
species_metrics <- species_table %>% 
  mutate(data = pmap(list(species, metric, resolution), ~get_species_metric(..1, ..2, ..3))) %>% 
  mutate(resolution = fct_relevel(resolution, c("hr", "mr", "lr"))) %>% 
  arrange(species, metric, resolution)

species_metrics
```

This unnests the data and recalculates abundance (`abundance_test`) and the difference between actual `abundance` and `abundance_test`.
```{r}
species_table_unnested <- species_metrics %>%
  unnest(data) %>% 
  select(species, resolution, date, month, x, y, metric_desc, value) %>% 
  pivot_wider(id_cols = c(species, resolution, date, month, x, y),
              names_from = metric_desc,
              values_from = value) %>% 
  select(species, resolution, date, month, x, y, count, occurrence, abundance) %>% 
  mutate(abundance_test = count * occurrence,
         diff = abundance - abundance_test)
```

Grouping by month to get to the county level changes the grain of the data so much that `abundance_test` undershoots `abundance` by 20%. This occurs at all resolutions.
```{r}
species_metrics %>%
  unnest(data) %>% 
  select(species, resolution, date, month, x, y, metric_desc, value) %>% 
  pivot_wider(id_cols = c(species, resolution, date, month, x, y),
              names_from = metric_desc,
              values_from = value) %>% 
  select(species, resolution, date, month, x, y, count, occurrence, abundance) %>% 
  group_by(species, month, resolution) %>% 
  summarize(occurrence = mean(occurrence, na.rm = T),
            count = mean(count, na.rm = T),
            abundance = mean(abundance, na.rm = T)) %>% 
  ungroup() %>% 
  mutate(abundance_test = count * occurrence,
         diff = abundance - abundance_test) %>% 
  ggplot(aes(abundance, abundance_test)) +
  geom_abline() +
  geom_point() +
  facet_wrap(~resolution) +
  tune::coord_obs_pred()
```

Totally un-aggregated, `abundance_test` closely resembles `abundance`, but degrades as resolution decreases.
```{r}
species_table_unnested %>% 
  select(abundance, abundance_test, resolution) %>% 
  drop_na() %>% 
  ggplot(aes(abundance, abundance_test)) +
  geom_density_2d_filled(contour_var = "ndensity") +
  facet_wrap(~resolution) +
  tune::coord_obs_pred() +
  coord_cartesian(xlim = c(0, 8),
                  ylim = c(0, 8)) +
  guides(fill = guide_colorsteps())
```

At lower resolutions, the difference is positively skewed, which means that `abundance` is higher than `abundance_test.`
```{r}
species_table_unnested %>% 
  drop_na(diff) %>% 
  ggplot(aes(diff)) +
  geom_histogram() +
  facet_wrap(~resolution, scale = "free_y", ncol = 1)
```

At the highest resolution, `diff` is heteroskedastic. At lower resolutions, there are patterns to the error. When `occurrence` is between 50% and 90%, `diff` is again skewed positively.
```{r}
species_table_unnested %>% 
  drop_na(occurrence, diff) %>% 
  ggplot(aes(occurrence, diff)) +
  geom_density_2d_filled(contour_var = "ndensity") +
  facet_wrap(~resolution) + 
  coord_cartesian(ylim = c(-.2, .2)) +
  scale_x_percent() +
  guides(fill = guide_colorsteps())
```

This was a useful exercise for me to understand how the geographic resolution and other aggregation of the data can affect estimated metrics, specifically in the citizen science context. 
