---
title: Graphing seasonality in eBird bird sightings
author: Conor Tompkins
date: '2020-03-31'
slug: graphing-seasonality-in-ebird-bird-sightings
categories:
  - R
tags: []
---

Over the winter I became interested in birding. Sitting in your back yard doing nothing but watching birds fly around is quite relaxing. Naturally I am looking for ways to optimize and quantify this relaxing activity. [eBird](https://ebird.org/home) lets you track your bird sightings and research which birds are common or more rare in your area. Luckily, the folks at [ROpenSci](https://ropensci.org/) have the `{rebird}` [package](https://cran.r-project.org/web/packages/rebird/index.html), which provides an easy interface to the eBird API.

In this post I will graph the seasonality of observation frequency of the top 10 birds in Pennsylvania. Frequency in this context is the % of eBird checklists that the bird appeared in during a given period.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE)

options(scipen = 99, digits = 2)
```

Load up packages:
```{r}
library(tidyverse)
library(lubridate)
library(vroom)
library(janitor)
library(rebird)
library(hrbrthemes)
library(ggrepel)
library(gganimate)

theme_set(theme_ipsum())
```

The `ebirdfreq` takes a location and time period and returns the frequency and sample size for the birds returned in the query.
```{r}
df_freq_raw <- ebirdfreq(loctype = 'states', loc = 'US-PA', startyear = 2019,
                         endyear = 2019, startmonth = 1, endmonth = 12)

df_freq_raw
```

This does some light data munging to get the data in shape.
```{r}
df_freq_clean <- df_freq_raw %>% 
  clean_names() %>%
  separate(month_qt, into = c("month", "week")) %>% 
  mutate(week = as.numeric(week),
         month = ymd(str_c("2019", month, "01", sep = "-")),
         month = month(month, label = TRUE, abbr = TRUE),
         state = "PA") %>% 
  rename(common_name = com_name) %>% 
  arrange(common_name, month, week)

df_freq_clean
```

This takes the month-week time series and summarizes to the month level:
```{r}
df_month <- df_freq_clean %>% 
  group_by(common_name, month) %>% 
  summarize(sample_size_mean = mean(sample_size),
            frequency_mean = mean(frequency) %>% round(2)) %>%
  ungroup()

df_month
```

Here I find the top 10 birds in terms of average monthly observation frequency:
```{r}
df_top_birds <- df_freq_clean %>% 
  group_by(common_name) %>% 
  summarize(sample_size_mean = mean(sample_size),
            frequency_mean = mean(frequency) %>% round(2)) %>% 
  ungroup() %>% 
  arrange(desc(frequency_mean)) %>% 
  select(common_name) %>% 
  slice(1:10)

df_top_birds
```

This basic line graph shows some of the pattern of seasonality, but fails to show the cyclical nature of the data.
```{r}
df_month %>% 
  semi_join(df_top_birds) %>% 
  ggplot(aes(month, frequency_mean, group = common_name)) +
    geom_line() +
    scale_y_percent() +
    labs(title = "Bird observation frequency",
         subtitle = "Top 10 birds in PA, 2019",
         x = NULL,
         y = "Mean frequency",
         caption = "Data from ebird.org. @conorotompkins")
```

I use `coord_polar` to change the coordinate system to match the cyclical flow of the months:
```{r}
df_month %>% 
  semi_join(df_top_birds) %>% 
  ggplot(aes(month, frequency_mean, group = common_name)) +
    geom_polygon(color = "black", fill = NA, size = .5) +
    coord_polar() +
    scale_y_percent() +
    labs(title = "Bird observation frequency",
         subtitle = "Top 10 birds in PA, 2019",
         x = NULL,
         y = "Mean frequency",
         caption = "Data from ebird.org. @conorotompkins")
```

`gganimate` lets me focus on one species at a time while showing all the data.
```{r}
plot_animated <- df_month %>% 
  semi_join(df_top_birds) %>% 
  mutate(common_name = fct_inorder(common_name)) %>% 
  ggplot(aes(month, frequency_mean)) +
  geom_polygon(data = df_month %>% rename(name = common_name),
               aes(group = name),
               color = "grey", fill = NA, size = .5) +
  geom_polygon(aes(group = common_name),
               color = "blue", fill = NA, size = 1.2) +
  coord_polar() +
  #facet_wrap(~common_name) +
  scale_y_percent() +
   labs(subtitle = "Most frequently observed birds in PA (2019)",
        x = NULL,
        y = "Frequency of observation",
        caption = "Data from ebird.org. @conorotompkins") +
  theme(plot.margin = margin(2, 2, 2, 2),
        plot.title = element_text(color = "blue"))

plot_animated +
  transition_manual(common_name) +
  ggtitle("{current_frame}")
```
