---
# Documentation: https://sourcethemes.com/academic/docs/managing-content/

title: "House Price Estimator Dashboard"
subtitle: ""
summary: "Use this interactive dashboard to model house prices in Allegheny County, PA"
authors: [Conor Tompkins]
tags: []
categories: []
date: 2021-01-28T15:23:43-05:00
lastmod: 2021-01-28T15:23:43-05:00
featured: true
draft: false

# Featured image
# To use, add an image named `featured.jpg/png` to your page's folder.
# Focal points: Smart, Center, TopLeft, Top, TopRight, Left, Right, BottomLeft, Bottom, BottomRight.
image:
  caption: ""
  focal_point: ""
  preview_only: true

# Projects (optional).
#   Associate this post with one or more of your projects.
#   Simply enter your project's folder or file name without extension.
#   E.g. `projects = ["internal-project"]` references `content/project/deep-learning/index.md`.
#   Otherwise, set `projects = []`.
projects: []
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      warning = FALSE, 
                      message = FALSE)
```
```{r include=FALSE}
library(tidyverse)
library(tidymodels)
library(hrbrthemes)

theme_set(theme_ipsum())

options(scipen = 999, digits = 4)
```

[Click here to view the full dashboard](https://conorotompkins.shinyapps.io/house_price_estimator/)

<iframe height="800" width="1000" frameborder="yes" src=" https://conorotompkins.shinyapps.io/house_price_estimator/"> </iframe>

# Model discussion

I trained 3 models against the assessment data:

* Linear model
* Random Forest
* Bagged Tree

I chose the Bagged Tree model because it performed about as well as the Random Forest model, but it predicts against new data much faster. Prediction speed is important because the UI for the dashboard has to update very quickly.

```{r include=FALSE}
total_records <- read_csv("data/bag_full_model_results.csv") %>% 
  nrow()
```

Total observations: `r comma(total_records)`

## Training set metrics (75% of total observations)
```{r include=FALSE}
train_metrics <- read_csv("data/train_metrics.csv")
```

```{r}
train_metrics %>% 
  select(model_name, dataset, .metric, .estimate) %>% 
  pivot_wider(names_from = .metric, values_from = .estimate)
```

```{r}
train_metrics %>% 
  select(model_name, .metric, .estimate) %>% 
  pivot_wider(names_from = .metric, values_from = .estimate) %>% 
  ggplot(aes(rmse, rsq, label = model_name)) +
  geom_label() +
  scale_x_continuous(label = dollar) +
  coord_cartesian(xlim = c(65000, 79000)) +
  labs(x = "Root Mean Squared Error",
       y = "R^2")
```

## Test set metrics (25% of total observations)
```{r include=FALSE}
test_metrics <- read_csv("data/test_metrics.csv")
```

```{r}
test_metrics %>% 
  select(model_name, dataset, .metric, .estimate) %>% 
  pivot_wider(names_from = .metric, values_from = .estimate)
```

```{r}
test_metrics %>% 
  select(model_name, .metric, .estimate) %>% 
  pivot_wider(names_from = .metric, values_from = .estimate) %>% 
  ggplot(aes(rmse, rsq, label = model_name)) +
  geom_label() +
  scale_x_continuous(label = dollar) +
  coord_cartesian(xlim = c(65000, 83000)) +
  labs(x = "Root Mean Squared Error",
       y = "R^2")
```

```{r include=FALSE}
full_model_results <- read_csv("data/bag_full_model_results.csv")
```

## Observed vs. Predicted (full dataset)

```{r}
full_model_results %>% 
  ggplot(aes(sale_price_adj, 10^.pred)) +
  geom_density_2d_filled() +
  geom_abline(lty = 2, color = "white") +
  scale_x_log10(label = dollar) +
  scale_y_log10(label = dollar) +
  coord_cartesian(xlim = c(30000, 10^6),
                  ylim = c(30000, 10^6)) +
  labs(x = "Actual sale price",
       y = "Predicted sale price",
       title = "Observed vs. Predicted sale price",
       fill = "Density of observations")
```

## Variable Importance
```{r include=FALSE}
var_imp <- read_csv("data/bagged_tree_variable_importance.csv")
```

```{r fig.height=14}
var_imp %>% 
  mutate(term = fct_reorder(term, value)) %>% 
  ggplot(aes(value, term)) +
  geom_point() +
  scale_x_comma() +
  labs(x = "Importance",
       y = NULL)
```
