---
# Documentation: https://sourcethemes.com/academic/docs/managing-content/

title: 'Comparing Healthy Ride Usage Pre And "Post" COVID-19'
subtitle: ""
summary: ""
authors: [Conor Tompkins]
tags: [R, Pittsburgh, Healthy Ride]
categories: [R, Pittsburgh, Healthy Ride]
date: 2020-08-13T13:54:25-04:00
lastmod: 2020-08-13T13:54:25-04:00
featured: false
draft: false

# Featured image
# To use, add an image named `featured.jpg/png` to your page's folder.
# Focal points: Smart, Center, TopLeft, Top, TopRight, Left, Right, BottomLeft, Bottom, BottomRight.
image:
  caption: ""
  focal_point: ""
  preview_only: false

# Projects (optional).
#   Associate this post with one or more of your projects.
#   Simply enter your project's folder or file name without extension.
#   E.g. `projects = ["internal-project"]` references `content/project/deep-learning/index.md`.
#   Otherwise, set `projects = []`.
projects: []
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE)
```

Lawrence Andrews asked me on Twitter if there had been a change in Health Ride usage after COVID-19.
<blockquote class="twitter-tweet"><p lang="en" dir="ltr">Would be interested to see this <a href="https://twitter.com/healthyridepgh?ref_src=twsrc%5Etfw">@healthyridepgh</a> data to compare pre-covid (2019) and during (2020)</p>&mdash; Lawrence Andrews (@lawrenceandrews) <a href="https://twitter.com/lawrenceandrews/status/1293879786528546816?ref_src=twsrc%5Etfw">August 13, 2020</a></blockquote> <script async src="https://platform.twitter.com/widgets.js" charset="utf-8"></script> 

The `{tidyverts}` universe of packages from Rob Hyndman provides a lot of tools that let you interrogate time series data. I will use some of these tools to decompose the Healthy Ride time series and see if there was a change.
```{r}
library(tidyverse)
library(lubridate)
library(janitor)
library(tsibble)
library(feasts)
library(hrbrthemes)

options(scipen = 999, digits = 4)

theme_set(theme_ipsum(base_size = 20, 
                      strip_text_size = 18, 
                      axis_title_size = 18))
```

I had already combined the usage data from the [WPRDC](https://data.wprdc.org/dataset/healthyride-trip-data) with `list.files` and `map_df(read_csv)`, so I can just read in the combined CSV file:
```{r}
data <- read_csv("data/combined_ride_data.csv")
```

Summarizing the number of rides per day shows that the data is very seasonal. The red line is on March 6, which is the date of the first *known* positive COVID-19 case in the state.
```{r}
data %>% 
  count(date, name = "number_of_rides", sort = TRUE) %>% 
  filter(!is.na(date)) %>% 
  ggplot(aes(date, number_of_rides)) +
  geom_point(alpha = .5, size = .5) +
  geom_vline(xintercept = ymd("2020-03-06"), color = "red")
```
I use the `{tsibble}` package to make a time series tibble and fill in a few gaps in the data. Then I create 3 different models to decompose the time series. I will compare these 3 models to see which strips away the seasonality the best.
```{r}
dcmp <- data %>%
  mutate(time = date) %>% 
  count(time, name = "number_of_rides") %>% 
  as_tsibble(index = time) %>% 
  tsibble::fill_gaps(number_of_rides = 0) %>% 
  model(STL(number_of_rides),
        STL(number_of_rides ~ season(window = Inf)),
        STL(number_of_rides ~ trend(window=7) + season(window='periodic'),
            robust = TRUE))

components(dcmp) %>% 
  glimpse()
```

This code pivots the data long and plots the true number of rides per day and the estimate of the underlying trend per model. The "season_adjust" panel shows the number of rides adjusted for seasonal effects, the "trend" panel shows the underlying trend, and the "remainder" panel shows how much the seasonal adjustment missed by.
```{r fig.height=10, fig.width=10}
components(dcmp) %>% 
  pivot_longer(cols = number_of_rides:season_adjust) %>% 
  mutate(name = factor(name, levels = c("number_of_rides", "season_adjust",
                                        "trend", "seasonal",
                                        "season_year", "season_week",
                                        "random", "remainder"))) %>% 
  filter(!is.na(value)) %>% 
  filter(name == "trend" | name == "season_adjust" | name == "number_of_rides" | name == "remainder") %>% 
  ggplot(aes(time, value, color = .model)) +
  geom_point(alpha = .6, size = .6) +
  annotate(geom = "rect", 
           xmin = ymd("2020-03-06"), xmax = ymd("2020-12-31"),
           ymin = -Inf, ymax = Inf, 
           fill = "red", alpha = .1) +
  facet_grid(name ~ .model, scales = "free_y", labeller = label_wrap_gen()) +
  guides(color = FALSE)
```

I am not a time series expert, but it appears that the most basic STL model `STL(number_of_rides)` does the best job because that model's "trend" panel shows the least seasonality.
```{r fig.height=10, fig.width=10}
components(dcmp) %>% 
  pivot_longer(cols = number_of_rides:season_adjust) %>% 
  mutate(name = factor(name, levels = c("number_of_rides", "season_adjust",
                                        "trend", "seasonal",
                                        "season_year", "season_week",
                                        "random", "remainder"))) %>% 
  filter(!is.na(value)) %>% 
  filter(name == "trend" | name == "season_adjust" | name == "number_of_rides" | name == "remainder") %>% 
  filter(.model == "STL(number_of_rides)") %>% 
  ggplot(aes(time, value, color = .model)) +
  geom_point(alpha = .6, size = .6, color = "#619CFF") +
  annotate(geom = "rect", 
           xmin = ymd("2016-03-06"), xmax = ymd("2016-03-30"),
           ymin = -Inf, ymax = Inf, 
           fill = "black", alpha = .3) +
  annotate(geom = "rect", 
           xmin = ymd("2017-03-06"), xmax = ymd("2017-03-30"),
           ymin = -Inf, ymax = Inf, 
           fill = "black", alpha = .3) +
  annotate(geom = "rect", 
           xmin = ymd("2018-03-06"), xmax = ymd("2018-03-30"),
           ymin = -Inf, ymax = Inf, 
           fill = "black", alpha = .3) +
  annotate(geom = "rect", 
           xmin = ymd("2019-03-06"), xmax = ymd("2019-03-30"),
           ymin = -Inf, ymax = Inf, 
           fill = "black", alpha = .3) +
  annotate(geom = "rect", 
           xmin = ymd("2020-03-06"), xmax = ymd("2020-03-06") + 60,
           ymin = -Inf, ymax = Inf, 
           fill = "red", alpha = .3) +
  facet_grid(name ~ .model, scales = "free_y", labeller = label_wrap_gen()) +
  guides(color = FALSE)
```

Focusing on that model, it appears that the trend dropped in mid-March, but rebounded to normal levels quickly. I highlighted the data from previous Marches to see if there was a recurring dip in March.